---
title: " Home Assignment 1<br><font color='#F48225' size='5'> 🍷 Wine production – RNAseq of the Flor yeast (Course 2025/2026)</font>"
author: "<font color='#A8A8A8' size='3'>Alicia Mañas, Lídia Sánchez and Paula Artiz<br>Take a look in GitHub: https://github.com/aliciaamanaas/DV-home-assigment/tree/main</font>"
date: "<font color='#A8A8A8' size='3'>Bachelor's Degree in Bioinformatics - May 2026</font>"

output:
  html_document:
    theme: yeti
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    css: https://cdnjs.cloudflare.com/ajax/libs/font-awesome/5.11.2/css/all.css
    self_contained: yes
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 4, fig.align = "center",
                      warning = FALSE, message = FALSE)
```

```{=html}
<style>
  @import url(https://fonts.googleapis.com/css?family=Fira+Sans:300,300i,400,400i,500,500i,700,700i);
  @import url(https://cdn.rawgit.com/tonsky/FiraCode/1.204/distr/fira_code.css);

  body { font-family: 'Fira Sans', sans-serif; }

  .code_box {
    padding: 1em;
    background: white;
    color: #222222;
    border: 2px solid #222222;
    border-radius: 10px;
    margin-bottom: 10px;
  }

  .alternative_box {
    padding: 1em;
    background: #fff8f8;
    border: 2px solid red;
    border-radius: 10px;
    margin-bottom: 10px;
  }

  .question_box {
    padding: 1em;
    background: #f4f8ff;
    border: 2px solid #3366cc;
    border-radius: 10px;
    margin-bottom: 10px;
  }

  .answer_box {
    padding: 1em;
    background: #f0fff4;
    border: 2px solid #28a745;
    border-radius: 10px;
    margin-bottom: 10px;
  }

  pre code { font-family: 'Fira Code', monospace; font-size: 0.9em; }
  h2 { border-bottom: 2px solid #F48225; padding-bottom: 4px; }
  h3 { color: #333; }
</style>
```

## Introduction

Flor yeasts (*Saccharomyces cerevisiae*) are the main microorganisms involved in the biological aging of sherry wines, where they form a velum on the surface of fortified must and enable survival under highly stressful winemaking conditions. This biological aging process contributes to key technological properties of sherry wines.

In this study, we analyze two industrial flor yeast strains used for the production of sherry wines A and B, which differ in flavor, alcohol percentage, and color. The goal is to determine whether these differences are associated with genetic or transcriptional variation between strains, or with other biochemical processes occurring during wine production.

RNAseq data were generated across six stages of velum development with four biological replicates per condition, resulting in 48 samples. After standard quality control, mapping, and gene quantification, the resulting count matrix is used to explore global expression patterns.

The main objective is to apply dimensionality reduction techniques to assess whether the two yeast strains show distinct transcriptomic profiles that could explain differences between the wine products.

---

## Part 1 — Load, Adapt the Data and Create Metadata

::: {.code_box data-latex=""}
**CODE #1: Load, adapt the data and create metadata**

Load the counts from the four replicate CSV files. You should get:

-   `meta_data`: a data frame with sample ID, wine type, time point and replicate.
-   `exprs_data`: a matrix with one row per sample and one column per gene.

Check data is correctly loaded: verify dimensions and confirm samples are
in the same order across both objects.
:::

### Install required packages

```{r install_packages, eval=FALSE}
# Install required packages (only needed once per session)
if (!requireNamespace("gtsummary", quietly = TRUE))
  install.packages("gtsummary")
if (!requireNamespace("stringr", quietly = TRUE))
  install.packages("stringr")
```

### Load data from separate CSV files

```{r load_data, cache=TRUE}
library(stringr)

setwd("/Users/paulaartizduenas/Desktop/DV - Group Project/data") # Should be changed to personal directory.

rep_files <- list(
  rep1 = "rep1.csv",
  rep2 = "rep2.csv",
  rep3 = "rep3.csv",
  rep4 = "rep4.csv"
)

# Read each file: genes x samples
rep_list <- lapply(rep_files, function(f) {
  m <- read.csv(f, row.names = 1, check.names = FALSE)
  as.matrix(m)
})

# Merge all replicates column-wise into one genes x samples matrix
exprs_data <- do.call(cbind, rep_list)
```

### Transpose and check dimensions

```{r transpose_check}
# Transpose: rows = samples, columns = genes
exprs_data <- t(exprs_data)

print(class(exprs_data))
print(dim(exprs_data)) # expected: 48 x 7126
print(exprs_data[1:5, 1:6])
```

### Create metadata

```{r create_metadata}
sample_names <- rownames(exprs_data)

library(stringr) # for str_match()

meta_data <- data.frame(
  row.names      = sample_names,
  wine           = str_match(sample_names, "^([AB])\\.")[, 2],
  time           = as.integer(str_match(sample_names, "\\.t(\\d+)_")[, 2]),
  replicate      = as.integer(str_match(sample_names, "_rep(\\d+)$")[, 2]),
  stringsAsFactors = FALSE
)

print(class(meta_data))
print(dim(meta_data)) # expected: 48 x 3
print(head(meta_data))
```

### Check samples are in the same order

```{r check_order}
table(rownames(meta_data) == rownames(exprs_data))
```

---

::: {.question_box data-latex=""}
**Q1. How many samples belong to each wine type and to each replicate?**
:::

```{r q1_summary, message=FALSE}
library(gtsummary)

meta_data |>
  gtsummary::tbl_summary(by = "wine") |>
  add_overall() |>
  add_n() |>
  bold_labels()
```

::: {.answer_box data-latex=""}
**ANSWER Q1:**

The dataset contains **48 samples** in total: **24 from wine A** and **24 from wine B**.
Each wine was measured at 6 velum development time points (t1–t6) with
4 biological replicates per time point, giving 8 samples per time point
across both wines.

The `wine` variable plays the same role as *treatment* in the teacher's
dataset, and `replicate` plays the role of *experiment*.
:::

::: {.alternative_box data-latex=""}
**CODE ALTERNATIVES:** You can inspect the sample distribution with base
R functions instead of `gtsummary`:

```{r q1_alt, eval=FALSE}
# Simple cross-tabulation: replicate x wine
table(meta_data$replicate, meta_data$wine)

# Or a quick summary
summary(meta_data)
str(meta_data)
```
:::

---

## Part 2 — PCA Representation

::: {.code_box data-latex=""}
**CODE #2: PCA representation**

Represent the data with a PCA projection using non-scaled, scaled and
normalised data. Scale or normalise data appropriately. Show the
different steps and comparison with raw data. Argue whether batch effects
are present between the replicates. Remove outliers if necessary.
:::

### Setup: load additional libraries

```{r setup_pca, message=FALSE}
library(ggplot2)
library(ggfortify)
library(factoextra)
library(dplyr)
library(ggpubr)

# Re-encode replicate as factor for plotting
meta_data$replicate <- as.factor(meta_data$replicate)
```

---

### PCA #1 — Not Scaled (`scale = FALSE`) {.tabset}

```{r pca_unscaled, cache=TRUE}
pca_res <- stats::prcomp(exprs_data, scale = FALSE)
```

#### Wine type — factoextra

```{r pca1_wine_fviz}
factoextra::fviz_pca_ind(pca_res,
                         geom.ind = "point",
                         col.ind = meta_data$wine,
                         addEllipses = TRUE,
                         legend.title = "Wine")
```

#### Wine type — autoplot

```{r pca1_wine_auto}
ggplot2::autoplot(pca_res, data = meta_data, colour = "wine") +
  theme_classic()
```

#### Wine type — base plot

```{r pca1_wine_base}
plot(pca_res$x, pch = 19, col = as.factor(meta_data$wine))
legend("topright", legend = levels(as.factor(meta_data$wine)),
       col = 1:2, pch = 19)
```

#### Replicate — factoextra

```{r pca1_rep_fviz}
factoextra::fviz_pca_ind(pca_res,
                         geom.ind    = "point",
                         col.ind     = meta_data$replicate,
                         addEllipses = TRUE,
                         legend.title = "Replicate")
```

#### Replicate — autoplot

```{r pca1_rep_auto}
ggplot2::autoplot(pca_res, data = meta_data, colour = "replicate") +
  theme_classic()
```

#### Replicate — base plot

```{r pca1_rep_base}
plot(pca_res$x, pch = 19, col = as.integer(meta_data$replicate))
legend("topright", legend = levels(meta_data$replicate),
       col = 1:4, pch = 19)
```

#### Variance explained — factoextra

```{r pca1_eig_fviz}
factoextra::fviz_eig(pca_res,
                     main = "Principal components: Variance explained")
```

#### Variance explained — base plot

```{r pca1_eig_base}
plot(pca_res)
```

#### {.unnumbered}

---

::: {.question_box data-latex=""}
**Q2. According to the PCA plot, explain how samples are distributed.
Justify whether you think that there is any batch effect in our samples
or not associated to the replicate ID.**
:::

::: {.answer_box data-latex=""}
**ANSWER Q2:**

Most of the variance does not correlate with wine type. However, there
is a clear clustering by replicate: **PC2 clearly separates samples from
different replicates**.

In summary, there is a noticeable **batch effect** caused by the
replicates. The differences caused by technical aspects between
replicates hide the important biological difference (wine A vs. wine B)
and need to be corrected through normalization.
:::

---

### PCA #2 — Scaled (`scale = TRUE`) {.tabset}

```{r pca_scaled, cache=TRUE}
print(dim(exprs_data))
exprs_data_clean <- exprs_data[, colSums(exprs_data) > 0]
print(dim(exprs_data_clean))
pca_res_scaled <- stats::prcomp(exprs_data_clean, scale = TRUE)

factoextra::fviz_eig(pca_res_scaled,
                     main = "Principal components: Variance explained — scaled")
```

#### Wine type — factoextra

```{r pca2_wine_fviz}
factoextra::fviz_pca_ind(pca_res_scaled,
                         geom.ind    = "point",
                         col.ind     = meta_data$wine,
                         addEllipses = TRUE,
                         legend.title = "Wine")
```

#### Wine type — autoplot

```{r pca2_wine_auto}
ggplot2::autoplot(pca_res_scaled, data = meta_data, colour = "wine") +
  theme_classic()
```

#### Replicate — factoextra

```{r pca2_rep_fviz}
factoextra::fviz_pca_ind(pca_res_scaled,
                         geom.ind    = "point",
                         col.ind     = meta_data$replicate,
                         addEllipses = TRUE,
                         legend.title = "Replicate")
```

#### Replicate — autoplot

```{r pca2_rep_auto}
ggplot2::autoplot(pca_res_scaled, data = meta_data, colour = "replicate") +
  theme_classic()
```

#### Wine + Replicate — factoextra

```{r pca2_both_fviz}
factoextra::fviz_pca_ind(pca_res_scaled,
                         geom.ind = "point",
                         col.ind  = meta_data$replicate,
                         addEllipses = TRUE) +
  geom_point(aes(shape = meta_data$wine))
```

#### Wine + Replicate — autoplot

```{r pca2_both_auto}
ggplot2::autoplot(pca_res_scaled,
                  data   = meta_data,
                  colour = "replicate",
                  shape  = "wine") +
  theme_classic() +
  scale_shape_manual(values = c(3, 17))
```

#### {.unnumbered}

---

::: {.question_box data-latex=""}
**Q3. Has the batch effect been corrected after scaling? If not, is
there a clear distinction between batches in PC1 or PC2?**
:::

::: {.answer_box data-latex=""}
**ANSWER Q3:**

Scaling by itself **does not eliminate the batch effect**. Replicates
continue to group apart, and now PC1 somewhat reflects the batch effect.
Scaling balances the gene inputs but cannot fix variations in total
library size across samples — direct normalization is required.
:::

---

### CODE #4 — Understanding the Batch Effect

```{r batch_effect_data}
head(rownames(exprs_data_clean))
head(rowSums(exprs_data_clean))

count_df <- data.frame(row.names = rownames(exprs_data_clean),
                       total_exp = rowSums(exprs_data_clean))
meta_data_all <- merge(count_df, meta_data, by = "row.names")
meta_data_all$Row.names <- as.factor(meta_data_all$Row.names)
```

#### Total expression bar — replicate

```{r batch_bar_rep}
ggplot(data = meta_data_all, aes(x = Row.names, y = total_exp, fill = replicate)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

#### Total expression bar — wine type

```{r batch_bar_wine}
ggplot(data = meta_data_all, aes(x = Row.names, y = total_exp, fill = wine)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

#### Total expression bar — gradient

```{r batch_bar_gradient}
ggplot(data = meta_data_all,
       aes(x = Row.names, y = total_exp, color = total_exp)) +
  geom_bar(stat = "identity") +
  theme_classic() +
  scale_color_gradient(low = "green", high = "red") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

#### Boxplot by replicate

```{r batch_box_rep}
ggpubr::ggboxplot(meta_data_all, x = "replicate",
                  y       = "total_exp",
                  fill    = "replicate",
                  add     = "jitter",
                  palette = "jama") +
  stat_compare_means(method = "wilcox.test") +
  labs(title = "Replicate")
```

#### Boxplot by wine type

```{r batch_box_wine}
ggpubr::ggboxplot(meta_data_all, x = "wine",
                  y       = "total_exp",
                  fill    = "wine",
                  add     = "jitter",
                  palette = "jama") +
  stat_compare_means() +
  labs(title = "Wine type")
```

#### Boxplot — both (faceted by replicate)

```{r batch_box_facet}
ggpubr::ggboxplot(meta_data_all, x = "wine",
                  y       = "total_exp",
                  fill    = "wine",
                  add     = "jitter",
                  palette = "jama") +
  stat_compare_means() +
  labs(title = "Wine type") +
  facet_grid(~replicate)
```

::: {.alternative_box data-latex=""}
**CODE ALTERNATIVES:** Base plot to inspect total counts per sample.

```{r batch_base, eval=TRUE}
plot(rowSums(exprs_data_clean), type = "h", col = as.integer(meta_data$replicate),
     ylab = "Total counts", xlab = "Sample")
plot(rowSums(exprs_data_clean), type = "h", col = as.factor(meta_data$wine),
     ylab = "Total counts", xlab = "Sample")
```
:::

```{r pca_expr_colour}
meta_data["Expression"] <- rowSums(exprs_data_clean)

ggplot2::autoplot(pca_res_scaled, data = meta_data, colour = "Expression") +
  theme_classic() +
  scale_color_gradient(low = "green", high = "red")
```

---

::: {.question_box data-latex=""}
**Q4. Is there any difference between replicates? Is there any
difference inside the same wine group? Do you expect samples with a high
value of total expression to have a positive or negative PC1?**
:::

```{r q4_lm, cache=TRUE}
pca_points <- dplyr::as_tibble(pca_res_scaled$x) |>
                bind_cols(meta_data) |>
                as.data.frame()
pca_points$exprs <- rowSums(exprs_data_clean)

pc1_mod <- lm(PC1 ~ exprs, pca_points)
sm      <- summary(pc1_mod)
print(sm)

r2   <- sm$r.squared
pval <- sm$coefficients[, 4][2]

color4points <- if (r2 > 0.6 & pval < 0.05) "red" else if (pval < 0.05) "orange" else "blue"

ggpubr::ggscatter(pca_points, x = "PC1", y = "exprs",
                  add      = "reg.line",
                  conf.int = TRUE,
                  cor.coef = TRUE,
                  color    = color4points,
                  alpha    = 0.6,
                  ggtheme  = theme_bw(),
                  xlab     = "PC1", ylab = "Total expression")
```

::: {.alternative_box data-latex=""}
**CODE ALTERNATIVES:**

```{r q4_alt, eval=TRUE}
plot(rowSums(exprs_data_clean), pca_res_scaled$x[, 1],
     xlab = "Total expression", ylab = "PC1",
     pch  = 19, col = as.integer(meta_data$replicate))
abline(lm(pca_res_scaled$x[, 1] ~ rowSums(exprs_data_clean)), col = "red")
legend("topright", legend = levels(meta_data$replicate),
       col = 1:4, pch = 19)
```
:::

```{r pca_expr_shape}
# Expression Gradient

ggplot2::autoplot(pca_res_scaled,
                  data   = meta_data,
                  colour = "Expression",
                  shape  = "replicate") +
  theme_classic() +
  scale_color_gradient(low = "green", high = "red") +
  scale_shape_manual(values = c(3, 17, 15, 18))
```

::: {.answer_box data-latex=""}
**ANSWER Q4:**

Replicates show significant differences in overall expression, which
confirms that **variations in library size are causing the batch
effect**. There is no significant difference in overall expression when
comparing wine A and wine B within the same replicate.

Samples with **high total expression are found on the left side of PC1**
(negative values), while those with low expression are on the right
side. Thus, PC1 is mainly driven by library size rather than the
biological differences between wine types.
:::

---

## Part 3 — Normalised Data

::: {.code_box data-latex=""}
**CODE #5 — Normalised data**

After this exploration we normalise each sample by its total expression
to remove the library-size effect, then re-run PCA with `scale = TRUE`.
:::

```{r norm_data, cache=TRUE}
exprs_data_norm <- exprs_data_clean / rowSums(exprs_data_clean)
pca_norm<- stats::prcomp(exprs_data_norm, scale = TRUE)
```

### PCA Normalised {.tabset}

#### Replicate — autoplot

```{r pca_norm_rep}
ggplot2::autoplot(pca_norm, data = meta_data, colour = "replicate") +
  theme_classic()
```

#### Wine type — autoplot

```{r pca_norm_wine}
ggplot2::autoplot(pca_norm, data = meta_data, colour = "wine") +
  theme_classic()
```

#### Both (colour: replicate, shape: wine)

```{r pca_norm_both}
ggplot2::autoplot(pca_norm,
                  data   = meta_data,
                  colour = "replicate",
                  shape  = "wine") +
  theme_classic() +
  scale_shape_manual(values = c(3, 17))
```

#### Colour: expression, shape: wine

```{r pca_norm_expr}
ggplot2::autoplot(pca_norm,
                  data   = meta_data,
                  colour = "Expression",
                  shape  = "wine") +
  theme_classic() +
  scale_color_gradient(low = "green", high = "red") +
  scale_shape_manual(values = c(3, 17))
```

#### Faceted by replicate

```{r pca_norm_facet}
ggplot2::autoplot(pca_norm,
                  data   = meta_data,
                  colour = "Expression",
                  shape  = "wine") +
  theme_classic() +
  scale_color_gradient(low = "green", high = "red") +
  facet_grid(~replicate) +
  scale_shape_manual(values = c(3, 17))
```

#### {.unnumbered}

::: {.alternative_box data-latex=""}
**CODE ALTERNATIVES:** You could also use base plot.

```{r pca_norm_base, eval=FALSE}
plot(pca_norm$x, pch = 19, col = as.integer(meta_data$replicate),
     main = "Normalised PCA — colour: replicate")
plot(pca_norm$x, pch = 19, col = as.factor(meta_data$wine),
     main = "Normalised PCA — colour: wine")
```
:::

---

::: {.question_box data-latex=""}
**Q5. According to the resulting plots, explain what groups can be
distinguished and what the main variables represented by PC1 and PC2 are
and the variance explained by each.**
:::

::: {.answer_box data-latex=""}
**ANSWER Q5:**

Once the data is properly normalized, the **batch effect completely
vanishes**, allowing samples from all four replicates to blend in the
PCA space.

Two distinct biological groups can be identified:

-   **PC1** distinguishes between the two wine types (A and B),
    representing the primary transcriptional variation between the yeast
    strains.
-   **PC2** appears to follow the velum development stages (t1–t6),
    showing how gene expression changes across the biological ageing
    periods.

This confirms that the initial batch effect was solely due to differences
in library sizes between replicates, and that the genuine biological
signals are revealed after normalization.
:::

---

## Part 4 — tSNE Representation

::: {.code_box data-latex=""}
**CODE #3: tSNE representation**

Represent the data with a tSNE projection using **raw data** and
**normalised data**. Scale or normalise data appropriately; show the
different steps and compare raw vs. normalised. Argue whether batch
effects are present between the replicates. Remove outliers if
necessary. Identify and show any correlation of the axes with expression
or any other metadata variable if possible.
:::

### Setup: load Rtsne library and prepare clean data

```{r setup_tsne, message=FALSE}
library(Rtsne)

# Remove all-zero columns and normalise (reuse exprs_data_clean / exprs_data_norm from Part 2-3)
exprs_clean <- exprs_data[, colSums(exprs_data) > 0]
exprs_norm  <- exprs_clean / rowSums(exprs_clean)
meta_data$Expression <- rowSums(exprs_clean)

cat("exprs_clean dims:", dim(exprs_clean), "\n")
cat("exprs_norm  dims:", dim(exprs_norm),  "\n")
```

---

### CODE #3.1 — tSNE on raw (non-normalised) data {.tabset}

```{r tsne_raw, cache=TRUE}
set.seed(42)
tsne_raw <- Rtsne(exprs_clean, dims = 2, perplexity = 10, verbose = FALSE)

df_raw <- data.frame(
  tSNE1      = tsne_raw$Y[, 1],
  tSNE2      = tsne_raw$Y[, 2],
  wine       = meta_data$wine,
  replicate  = meta_data$replicate,
  Expression = meta_data$Expression,
  time       = as.factor(meta_data$time)
)
```

#### Wine type

```{r tsne_raw_wine}
ggplot(df_raw, aes(x = tSNE1, y = tSNE2, colour = wine)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_colour_manual(values = c("A" = "#E74C3C", "B" = "#3498DB"),
                      labels = c("A" = "Wine A", "B" = "Wine B")) +
  labs(title  = "tSNE — Raw data | Colour: Wine type",
       colour = "Wine type") +
  theme_bw()
```

#### Replicate

```{r tsne_raw_rep}
ggplot(df_raw, aes(x = tSNE1, y = tSNE2, colour = replicate)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_colour_brewer(palette = "Set1") +
  labs(title  = "tSNE — Raw data | Colour: Replicate",
       colour = "Replicate") +
  theme_bw()
```

#### Total expression

```{r tsne_raw_expr}
ggplot(df_raw, aes(x = tSNE1, y = tSNE2, colour = Expression)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_colour_viridis_c(option = "plasma") +
  labs(title  = "tSNE — Raw data | Colour: Total expression",
       colour = "Total counts") +
  theme_bw()
```

#### {.unnumbered}

---

::: {.question_box data-latex=""}
**Q3a. According to the tSNE plot on raw data, how are samples
distributed? Is there evidence of a batch effect associated with
replicates?**
:::

::: {.answer_box data-latex=""}
**ANSWER Q3a:**

In the raw tSNE plot, samples are not randomly distributed and instead
form distinct clusters. There is clear evidence of a **batch effect
associated with replicates**, as samples group primarily according to
replicate rather than biological condition. This indicates that technical
variation is dominating the structure of the data.
:::

---

### CODE #3.2 — tSNE on normalised data {.tabset}

```{r tsne_norm, cache=TRUE}
set.seed(42)
tsne_norm <- Rtsne(exprs_norm, dims = 2, perplexity = 10, verbose = FALSE)

df_norm <- data.frame(
  tSNE1      = tsne_norm$Y[, 1],
  tSNE2      = tsne_norm$Y[, 2],
  wine       = meta_data$wine,
  replicate  = meta_data$replicate,
  Expression = meta_data$Expression,
  time       = as.factor(meta_data$time)
)
```

#### Wine type

```{r tsne_norm_wine}
ggplot(df_norm, aes(x = tSNE1, y = tSNE2, colour = wine)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_colour_manual(values = c("A" = "#E74C3C", "B" = "#3498DB"),
                      labels = c("A" = "Wine A", "B" = "Wine B")) +
  labs(title  = "tSNE — Normalised data | Colour: Wine type",
       colour = "Wine type") +
  theme_bw()
```

#### Replicate

```{r tsne_norm_rep}
ggplot(df_norm, aes(x = tSNE1, y = tSNE2, colour = replicate)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_colour_brewer(palette = "Set1") +
  labs(title  = "tSNE — Normalised data | Colour: Replicate",
       colour = "Replicate") +
  theme_bw()
```

#### Total expression

```{r tsne_norm_expr}
ggplot(df_norm, aes(x = tSNE1, y = tSNE2, colour = Expression)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_colour_viridis_c(option = "plasma") +
  labs(title  = "tSNE — Normalised data | Colour: Total expression",
       colour = "Total counts") +
  theme_bw()
```

#### Time point

```{r tsne_norm_time}
ggplot(df_norm, aes(x = tSNE1, y = tSNE2, colour = time)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_colour_brewer(palette = "RdYlBu", direction = -1) +
  labs(title  = "tSNE — Normalised data | Colour: Time point",
       colour = "Time point") +
  theme_bw()
```

#### {.unnumbered}

::: {.alternative_box data-latex=""}
**CODE ALTERNATIVES:** You can also colour by both wine and replicate
simultaneously using shape aesthetics.

```{r tsne_norm_alt, eval=FALSE}
ggplot(df_norm, aes(x = tSNE1, y = tSNE2,
                    colour = wine, shape = replicate)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_colour_manual(values = c("A" = "#E74C3C", "B" = "#3498DB")) +
  labs(title = "tSNE — Normalised | Wine (colour) & Replicate (shape)") +
  theme_bw()
```
:::

---

::: {.question_box data-latex=""}
**Q3b. After normalisation, has the batch effect been corrected? What
groups can you distinguish? What variables seem to drive the two axes?**
:::

::: {.answer_box data-latex=""}
**ANSWER Q3b:**
Once normalized, the samples mainly group by wine type, showing that the main pattern in the dataset relates to natural differences instead of technical influences. Samples taken at different times stay fairly close within each wine type group, implying that time differences are less significant than the type of wine. 

The distinct division between Wine A and Wine B shows that normalization maintained important biological differences while minimizing unrelated technical variation. Although some residual technical variation may still be present, batch-related effects no longer appear to dominate the clustering pattern.
:::

---

## Part 4 — tSNE Parameters

::: {.code_box data-latex=""}
**CODE #4: tSNE parameters**

Test the effect of **reproducibility**, **perplexity**, and
**iterations** on the tSNE result. Use the **normalised dataset only**.
Provide comments and thoughts about each parameter.
:::

---

### CODE #4.1 — Effect of reproducibility (random seed)

```{r tsne_seed, cache=TRUE}
# Run tSNE three times without a fixed seed to show non-determinism
tsne_res_1 <- Rtsne(exprs_norm, dims = 2, perplexity = 10)
tsne_res_2 <- Rtsne(exprs_norm, dims = 2, perplexity = 10)
tsne_res_3 <- Rtsne(exprs_norm, dims = 2, perplexity = 10)
```

#### Plot runs side by side

```{r tsne_seed_plot}
make_df <- function(res, run_label) {
  data.frame(x = res$Y[, 1], y = res$Y[, 2], run = run_label)
}

df_runs <- rbind(
  make_df(tsne_res_1, "Run 1"),
  make_df(tsne_res_2, "Run 2"),
  make_df(tsne_res_3, "Run 3")
)

ggplot(df_runs, aes(x = x, y = y)) +
  geom_point(alpha = 0.7) +
  facet_wrap(~run, nrow = 1) +
  labs(title = "tSNE — Effect of random seed (no set.seed)",
       x = "tSNE1", y = "tSNE2") +
  theme_bw()
```

---

::: {.question_box data-latex=""}
**Q4a. What do you observe when running tSNE multiple times without
fixing the seed? Why does this happen?**
:::

::: {.answer_box data-latex=""}
**ANSWER Q4a:**

We see that the **clusters remain broadly consistent** across different
tSNE runs, but their positions, orientations and relative distances
change each time. This happens because tSNE is a **stochastic method**:
it initialises coordinates randomly, so each run starts from a different
configuration and converges to a different (though equally valid) local
minimum. Fixing the seed with `set.seed()` makes runs fully
reproducible.
:::

---

### CODE #4.2 — Effect of perplexity

```{r tsne_perplexity, cache=TRUE}
# Try three different perplexity values (perplexity must be < nrow(data)/3)
set.seed(42)
tsne_perp5  <- Rtsne(exprs_norm, dims = 2, perplexity = 5)
tsne_perp10 <- Rtsne(exprs_norm, dims = 2, perplexity = 10)
tsne_perp15 <- Rtsne(exprs_norm, dims = 2, perplexity = 15)
```

#### Plot results for each perplexity value

```{r tsne_perplexity_plot}
make_df_perp <- function(res, perp_label) {
  data.frame(x = res$Y[, 1], y = res$Y[, 2],
             Perplexity = perp_label,
             wine = meta_data$wine)
}

df_perp <- rbind(
  make_df_perp(tsne_perp5,  "Perplexity = 5"),
  make_df_perp(tsne_perp10, "Perplexity = 10"),
  make_df_perp(tsne_perp15, "Perplexity = 15")
)

ggplot(df_perp, aes(x = x, y = y, colour = wine)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_colour_manual(values = c("A" = "#E74C3C", "B" = "#3498DB")) +
  facet_wrap(~Perplexity, nrow = 1) +
  labs(title  = "tSNE — Effect of perplexity (normalised data)",
       x      = "tSNE1", y = "tSNE2",
       colour = "Wine type") +
  theme_bw()
```

---

::: {.question_box data-latex=""}
**Q4b. How does perplexity affect the resulting tSNE plot? Which value
gives the most interpretable result and why?**
:::

::: {.answer_box data-latex=""}
**ANSWER Q4b:**

Perplexity is a parameter that roughly controls the **effective number
of neighbours** each point considers. Its effect is:

- **Low perplexity (5):** The algorithm focuses on very local
  structure, producing many small, fragmented clusters with a lot of
  noise.
- **Intermediate perplexity (10–15):** A better balance between local
  and global structure is achieved; the two biological groups become
  more clearly separated.
- **The most interpretable result is obtained with perplexity = 15**,
  as it shows the two wine groups more clearly and reduces the noise
  visible at perplexity = 5. Values that are too high would risk
  merging distinct groups.
:::

---

### CODE #4.3 — Effect of number of iterations

```{r tsne_iterations, cache=TRUE}
# Try three different max_iter values
set.seed(123)
tsne_iter100  <- Rtsne(exprs_norm, dims = 2, perplexity = 15, max_iter = 100)
tsne_iter500  <- Rtsne(exprs_norm, dims = 2, perplexity = 15, max_iter = 500)
tsne_iter1000 <- Rtsne(exprs_norm, dims = 2, perplexity = 15, max_iter = 1000)
```

#### Plot results for each iteration count

```{r tsne_iterations_plot}
make_df_iter <- function(res, iter_label) {
  data.frame(x = res$Y[, 1], y = res$Y[, 2],
             Iterations = iter_label,
             wine = meta_data$wine)
}

df_iter <- rbind(
  make_df_iter(tsne_iter100,  "100 iterations"),
  make_df_iter(tsne_iter500,  "500 iterations"),
  make_df_iter(tsne_iter1000, "1000 iterations")
)

df_iter$Iterations <- factor(df_iter$Iterations,
                              levels = c("100 iterations",
                                         "500 iterations",
                                         "1000 iterations"))

ggplot(df_iter, aes(x = x, y = y, colour = wine)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_colour_manual(values = c("A" = "#E74C3C", "B" = "#3498DB")) +
  facet_wrap(~Iterations, nrow = 1) +
  labs(title  = "tSNE — Effect of iterations (normalised data, perplexity = 15)",
       x      = "tSNE1", y = "tSNE2",
       colour = "Wine type") +
  theme_bw()
```

::: {.alternative_box data-latex=""}
**CODE ALTERNATIVES:** You can also inspect convergence by comparing KL
divergence (reported when `verbose = TRUE`).

```{r tsne_iter_alt, eval=FALSE}
set.seed(123)
tsne_verbose <- Rtsne(exprs_norm, dims = 2, perplexity = 15,
                      max_iter = 1000, verbose = TRUE)
```
:::

---

::: {.question_box data-latex=""}
**Q4c. How does the number of iterations affect convergence and the
final layout? What is a reasonable minimum for this dataset?**
:::

::: {.answer_box data-latex=""}
**ANSWER Q4c:**

The number of iterations affects the algorithm's ability to converge
and find a stable solution:

- **100 iterations:** The algorithm has not converged yet; clusters are
  poorly defined and the layout appears messy.
- **500 iterations:** Well-defined clusters can already be observed,
  suggesting the algorithm has reached a reasonable solution.
- **1000 iterations:** The layout is similar to 500 but slightly more
  refined; the improvement over 500 is marginal.

A **reasonable minimum for this dataset is 500 iterations**, as it
already yields interpretable and stable clusters. The default of 1000
is a safe choice to ensure full convergence.
:::

---

## Part 5 — Final Interpretation

::: {.code_box data-latex=""}
**CODE #5: Final publication-quality PCA and tSNE**

Create final, publication-quality representations for both PCA and tSNE
using the normalised dataset.

- Summarise conclusions for PC1 and PC2.
- Compare PCA and tSNE: do you reach the same biological conclusions?
- If any outlier was identified, comment on how each method deals with it.
:::

---

### CODE #5.1 — Final PCA plot (normalised data) {.tabset}

```{r pca5_run, cache=TRUE}
# PCA on normalised data
pca_norm <- stats::prcomp(exprs_norm, scale = TRUE)

# % variance explained for axis labels
var_exp  <- summary(pca_norm)$importance[2, ] * 100
pc1_lab  <- paste0("PC1 (", round(var_exp[1], 2), "%)")
pc2_lab  <- paste0("PC2 (", round(var_exp[2], 2), "%)")
```

#### Wine type & replicate

```{r pca5_main}
autoplot(pca_norm,
         data    = meta_data,
         colour  = "wine",
         shape   = "replicate") +
  scale_colour_manual(values = c("A" = "#E74C3C", "B" = "#3498DB"),
                      labels = c("A" = "Wine A", "B" = "Wine B")) +
  scale_shape_manual(values = c(16, 17, 15, 3)) +
  labs(title  = "PCA — Normalised data",
       x      = pc1_lab,
       y      = pc2_lab,
       colour = "Wine type",
       shape  = "Replicate") +
  theme_classic()
```

#### Faceted by replicate

```{r pca5_facet}
autoplot(pca_norm,
         data   = meta_data,
         colour = "wine",
         shape  = "replicate") +
  scale_colour_manual(values = c("A" = "#E74C3C", "B" = "#3498DB")) +
  scale_shape_manual(values = c(16, 17, 15, 3)) +
  facet_wrap(~replicate) +
  labs(title  = "PCA faceted by replicate",
       x      = pc1_lab,
       y      = pc2_lab,
       colour = "Wine type",
       shape  = "Replicate") +
  theme_bw()
```

#### {.unnumbered}

---

::: {.question_box data-latex=""}
**Q5a. What does PC1 represent? What does PC2 represent?
How much variance does each explain?
Are the conclusions consistent with Questions 2–4?**
:::

::: {.answer_box data-latex=""}
**ANSWER Q5a:**

**PC1** separates wine types A and B, indicating that the main source of
variation is driven by differences between yeast strains. It explains
**69.14%** of total variance.

**PC2** captures variation associated with velum development stages,
reflecting temporal changes in gene expression during the aging process.
It explains **24.71%** of the variance.

Together, PC1 and PC2 account for **93.85%** of total variance, meaning
that reducing the data to two dimensions retains almost all relevant
information. These conclusions are fully consistent with those from
Questions 2–4.
:::

---

### CODE #5.2 — Final tSNE plot (normalised data) {.tabset}

```{r tsne5_run, cache=TRUE}
library(Rtsne)

set.seed(123)
tsne_final <- Rtsne(exprs_norm, dims = 2, perplexity = 15, max_iter = 1000)

df_final <- data.frame(
  tSNE1     = tsne_final$Y[, 1],
  tSNE2     = tsne_final$Y[, 2],
  wine      = meta_data$wine,
  replicate = meta_data$replicate,
  time      = meta_data$time,
  Expression = meta_data$Expression
)
```

#### Wine type

```{r tsne5_wine}
ggplot(df_final, aes(x = tSNE1, y = tSNE2, colour = wine)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_colour_manual(values = c("A" = "#E74C3C", "B" = "#3498DB"),
                      labels = c("A" = "Wine A", "B" = "Wine B")) +
  labs(title  = "t-SNE — Normalised data | Colour: Wine type",
       colour = "Wine type") +
  theme_bw()
```

#### Replicate

```{r tsne5_rep}
ggplot(df_final, aes(x = tSNE1, y = tSNE2, colour = replicate)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_colour_brewer(palette = "Set1") +
  labs(title  = "t-SNE — Normalised data | Colour: Replicate",
       colour = "Replicate") +
  theme_bw()
```

#### Time point

```{r tsne5_time}
ggplot(df_final, aes(x = tSNE1, y = tSNE2, colour = as.factor(time))) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_colour_brewer(palette = "RdYlBu", direction = -1) +
  labs(title  = "t-SNE — Normalised data | Colour: Time point",
       colour = "Time point") +
  theme_bw()
```

#### {.unnumbered}

---

::: {.question_box data-latex=""}
**Q5b. What groups are visible? Do the axes carry the same biological
meaning as in PCA? Do PCA and tSNE lead to the same conclusions?**
:::

::: {.answer_box data-latex=""}
**ANSWER Q5b:**

Two clearly separated clusters are visible, corresponding to **wine type
A and wine type B**. Replicates are distributed within each wine group,
confirming that technical variation does not drive the main separation.

The t-SNE axes do **not** carry the same biological meaning as PCA axes:
t-SNE distances are not directly interpretable in terms of variance
explained. Its role is to preserve local neighbourhood relationships, not
to maximise global variance.

Despite this, both methods reach the **same biological conclusion**: the
dominant source of variation in this dataset is wine type, driven by
differences between the yeast strains used.
:::

---

### CODE #5.3 — Outlier analysis {.tabset}

```{r outlier5_detect}
# Tukey criterion applied to PC1 and PC2 scores
pca_scores <- as.data.frame(pca_norm$x)

tukey_outlier <- function(x) {
  Q1      <- quantile(x, 0.25)
  Q3      <- quantile(x, 0.75)
  IQR_val <- Q3 - Q1
  x < (Q1 - 1.5 * IQR_val) | x > (Q3 + 1.5 * IQR_val)
}

outlier_pc1 <- tukey_outlier(pca_scores$PC1)
outlier_pc2 <- tukey_outlier(pca_scores$PC2)
is_outlier  <- outlier_pc1 | outlier_pc2

cat("Outliers detected:\n")
print(rownames(pca_scores)[is_outlier])

meta_data$outlier <- is_outlier
df_final$outlier  <- is_outlier
```

#### PCA — outlier highlighted

```{r outlier5_pca}
autoplot(pca_norm,
         data   = meta_data,
         colour = "wine",
         shape  = "outlier") +
  scale_colour_manual(values = c("A" = "#E74C3C", "B" = "#3498DB")) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 8),
                     labels = c(`FALSE` = "Normal", `TRUE` = "Outlier")) +
  labs(title  = "PCA — Outlier highlighted (Tukey criterion)",
       x      = pc1_lab,
       y      = pc2_lab,
       colour = "Wine type",
       shape  = "Outlier") +
  theme_classic()
```

#### t-SNE — outlier highlighted

```{r outlier5_tsne}
ggplot(df_final, aes(x = tSNE1, y = tSNE2, colour = wine, shape = outlier)) +
  geom_point(size = 2.5, alpha = 0.9) +
  scale_colour_manual(values = c("A" = "#E74C3C", "B" = "#3498DB")) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 8),
                     labels = c(`FALSE` = "Normal", `TRUE` = "Outlier")) +
  labs(title  = "t-SNE — Outlier highlighted (Tukey criterion)",
       colour = "Wine type",
       shape  = "Outlier") +
  theme_bw()
```

#### PCA without outlier

```{r outlier5_pca_clean, cache=TRUE}
exprs_norm_clean <- exprs_norm[!is_outlier, ]
meta_clean       <- meta_data[!is_outlier, ]

pca_no_out  <- stats::prcomp(exprs_norm_clean, scale = TRUE)
var_no_out  <- summary(pca_no_out)$importance[2, ] * 100
pc1_no_lab  <- paste0("PC1 (", round(var_no_out[1], 2), "%)")
pc2_no_lab  <- paste0("PC2 (", round(var_no_out[2], 2), "%)")

autoplot(pca_no_out,
         data   = meta_clean,
         colour = "wine",
         shape  = "replicate") +
  scale_colour_manual(values = c("A" = "#E74C3C", "B" = "#3498DB")) +
  scale_shape_manual(values = c(16, 17, 15, 3)) +
  labs(title  = "PCA — Without outlier (B.t1_rep3 removed)",
       x      = pc1_no_lab,
       y      = pc2_no_lab,
       colour = "Wine type",
       shape  = "Replicate") +
  theme_classic()
```

#### {.unnumbered}

---

::: {.question_box data-latex=""}
**Q5c. Was an outlier detected? How does each method (PCA vs tSNE) place
it relative to the rest of the data? Does removing it change the
interpretation?**
:::

::: {.answer_box data-latex=""}
**ANSWER Q5c:**

**Yes**, one outlier was detected: **B.t1_rep3** (Wine B, time point 1,
replicate 3). Applying the Tukey criterion to the PCA scores, this sample
obtained a PC1 score of **459.6**, far above the upper fence of **73.8**
(Q3 + 1.5·IQR), making it unambiguously anomalous.

In **PCA**, the outlier dominates PC1 entirely, pulling the axis toward
itself and compressing all other 47 samples into an unreadable cluster,
severely limiting interpretability.

In **t-SNE**, the outlier is placed in an isolated position separated
from both wine-type clusters. However, because t-SNE optimises local
neighbourhood relationships rather than global variance, the remaining
samples are far less distorted and the two groups remain visible.

Removing B.t1_rep3 does **not change the biological conclusion** — wine
type remains the dominant source of variation. However, its removal is
fully justified by the Tukey criterion and greatly improves the clarity
of the PCA plot.
:::

---

## Final Summary

::: {.answer_box data-latex=""}
**FINAL SUMMARY:**

Both PCA and t-SNE consistently show that the primary driver of variation
in this dataset is **wine type**: samples from wine A and wine B form two
clearly separated groups. PC1 (69.14%) captures this inter-wine
difference, while PC2 (24.71%) reflects temporal changes in gene
expression across velum development stages (t1–t6). Replicates are well
distributed within each group, indicating good technical reproducibility
and no strong batch effect after normalisation. One outlier was
identified — **B.t1_rep3** — which disproportionately distorts the PCA
but can be removed on the basis of the Tukey criterion without altering
the overall biological interpretation. In summary, the differences
between wine A and wine B are driven by the yeast strains used, and this
signal is robust across both dimensionality reduction methods.
:::